---
title: "Untitled"
output: html_document
date: "2024-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(nflfastR)
library(dplyr)
library(readr)

```


--- Comparing Years---

```{r}
dataEarlyYears <- load_pbp(2003:2013)
dataLaterYears <- load_pbp(2013:2023)

goforitEarly <-  dataEarlyYears %>% filter(down==4&(play_type=="run"|play_type=="pass"))
goforitLater <-  dataLaterYears %>% filter(down==4&(play_type=="run"|play_type=="pass"))

```

```{r}

plot(goforitEarly$ydstogo,goforitEarly$yards_gained)
cor(goforitEarly$ydstogo,goforitEarly$yards_gained)

unique(dataEarlyYears$play_type)

plot(goforitLater$ydstogo,goforitLater$yards_gained)
cor(goforitLater$ydstogo,goforitLater$yards_gained)

unique(dataLaterYears$play_type)

```

---Overall---

```{r}

goforitEarly %>% filter(ydstogo>=1&ydstogo<=10)
yardstogo_plot <- function(df){
  conversion <- c()
  df$ydstogo <- ifelse(df$ydstogo>10,11,df$ydstogo)
for (i in 1:11){
  tbl <- df %>% filter(ydstogo==i) 
  conversion[i] <- tbl$fourth_down_converted %>% mean()
}
Yardstogo <- 1:11
ggdfEarly <- data.frame(Yardstogo,conversion)
ggplot(ggdfEarly,aes(x=Yardstogo,y=conversion)) + geom_point(size=5,shape=22,fill="blue") + theme(plot.title=element_text("Conversion Rate based on Yards to Go"),panel.background=element_rect(fill="lightgrey"),panel.grid.major.x=element_line(colour="yellow"),panel.grid.minor.x=element_line(colour="yellow"))}
goforitEarly %>% filter(ydstogo==5)
yardstogo_plot(goforitEarly)



goforitLater %>% filter(ydstogo>=1&ydstogo<=10)
yardstogo_plot <- function(df){
  conversion <- c()
  df$ydstogo <- ifelse(df$ydstogo>10,11,df$ydstogo)
for (i in 1:11){
  tbl <- df %>% filter(ydstogo==i) 
  conversion[i] <- tbl$fourth_down_converted %>% mean()
}
Yardstogo <- 1:11
ggdfLater <- data.frame(Yardstogo,conversion)
ggplot(ggdfLater,aes(x=Yardstogo,y=conversion)) + geom_point(size=5,shape=22,fill="blue") + theme(plot.title=element_text("Conversion Rate based on Yards to Go"),panel.background=element_rect(fill="lightgrey"),panel.grid.major.x=element_line(colour="yellow"),panel.grid.minor.x=element_line(colour="yellow"))}
goforitLater %>% filter(ydstogo==5)
yardstogo_plot(goforitLater)

```

```{r}

forty_to_fiftyEarly <- goforitEarly %>% filter(40<=yardline_100) %>% filter(yardline_100<50)
forty_to_fiftyEarly
yardstogo_plot(forty_to_fiftyEarly)

thirty_to_fortyEarly <- goforitEarly %>% filter(30<=yardline_100) %>% filter(yardline_100<40)
yardstogo_plot(thirty_to_fortyEarly)


twenty_to_thirtyEarly <- goforitEarly %>% filter(20<=yardline_100) %>% filter(yardline_100<30)
yardstogo_plot(twenty_to_thirtyEarly)


ten_to_twentyEarly <- goforitEarly %>% filter(10<=yardline_100) %>% filter(yardline_100<20)
yardstogo_plot(ten_to_twentyEarly)



forty_to_fiftyLater <- goforitLater %>% filter(40<=yardline_100) %>% filter(yardline_100<50)
yardstogo_plot(forty_to_fiftyLater)

thirty_to_fortyLater <- goforitLater %>% filter(30<=yardline_100) %>% filter(yardline_100<40)
yardstogo_plot(thirty_to_fortyLater)


twenty_to_thirtyLater <- goforitLater %>% filter(20<=yardline_100) %>% filter(yardline_100<30)
yardstogo_plot(twenty_to_thirtyLater)


ten_to_twentyLater <- goforitLater %>% filter(10<=yardline_100) %>% filter(yardline_100<20)
yardstogo_plot(ten_to_twentyLater)

```

--------- Just some things I'm noticing -----------

Differences in observation numbers

10-20 Yard Line : 
    2003-2013 -> 562
    2013-2023 -> 624
    
20-30 Yard Line :
    2003-2013 -> 707
    2013-2023 -> 730
    
30-40 Yard Line :
    2003-2013 -> 1315
    2013-2023 -> 1297
    
40-50 Yard Line :
    2003-2013 -> 832
    2013-2023 -> 1160
    
* Focus on the 40-50 yard line? There seems to be the biggest 
differences between years there *

First, is there a difference in the number of times they go for it - Looking
at difference in average number of times they go for it in a week

```{r}


weekGoForItEarly = NULL

for (i in 1:21) {
  weekGoForItEarly[i] <- nrow(goforitEarly[goforitEarly$week == i, ] %>% filter(40<=yardline_100) %>% filter(yardline_100<50))
}

weekGoForItLater = NULL

for (i in 1:21) {
  weekGoForItLater[i] <- nrow(goforitLater[goforitLater$week == i, ] %>% filter(40<=yardline_100) %>% filter(yardline_100<50))
}

t.test(weekGoForItLater, weekGoForItEarly, var.equal=TRUE)

```



So there is a statistical difference in the number of attempts, but is there
a difference in the success rate of these attempts?

```{r}

succEarly <- goforitEarly[goforitEarly$ydstogo <= goforitEarly$yards_gained,]
succLater <- goforitLater[goforitLater$ydstogo <= goforitLater$yards_gained,]

overallSuccEarly <- nrow(succEarly) / nrow(goforitEarly)
overallSuccLater <- nrow(succLater) / nrow(goforitLater)

succWeekEarly = NULL

for (i in 1:21) {
  succWeekEarly[i] <- nrow(succEarly[succEarly$week == i,]) / nrow(goforitEarly[goforitEarly$week == i, ])
}

succWeekLater = NULL

for (i in 1:21) {
  succWeekLater[i] <- nrow(succLater[succLater$week == i,]) / 
nrow(goforitLater[goforitLater$week == i, ])
}


t.test(succWeekLater, succWeekEarly, var.equal = TRUE)

```

With alpha = 0.05, this also is statistically significant. Now, is there a big
difference in the timing of when going for a 4th down?

```{r}

ggplot(data=goforitEarly,
       mapping = aes(x=quarter_seconds_remaining)) +
  geom_histogram(binwidth = 18)+
  labs(title = "4th Down Attempts by Time Left in the Quarter : 2003-2013")

ggplot(data=goforitLater,
       mapping = aes(x=quarter_seconds_remaining)) +
  geom_histogram(binwidth = 18) +
  labs(title = "4th Down Attempts by Time Left in the Quarter : 2013-2023")

```

