---
title: "FieldPositionConversion4th"
output: md_document
date: "2024-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nflfastR)
library(dplyr)
library(readr)

data <- load_pbp(2013:2023)
dim(data)

fours <- data %>% filter(down==4&(play_type=="punt"|play_type=="run"|play_type=="pass"))
goforit <-  data %>% filter(down==4&(play_type=="run"|play_type=="pass"))
posplays <- goforit %>% filter(yards_gained>=0)

goforit %>% filter(ydstogo>=1&ydstogo<=10)

```
```{r}
winners1 <- goforit %>% filter(home_team==posteam&result>0)
winners2 <- goforit %>% filter(away_team==posteam&result<0)
winners <- rbind(winners1,winners2)
losers1 <- goforit %>% filter(home_team==posteam&result<0)
losers2 <- goforit %>% filter(away_team==posteam&result>0)
losers <- rbind(losers1,losers2)
```

```{r}
conversion_rates <- c()
 for (yds in 1:10) {
      tbl <- winners %>% filter(ydstogo == yds)
      conversion_rates[yds] <- mean(tbl$fourth_down_converted, na.rm = TRUE)
 }
conversion_rates1 <- c()
for (yds in 1:10) {
      tbl <- losers %>% filter(ydstogo == yds)
      conversion_rates1[yds] <- mean(tbl$fourth_down_converted, na.rm = TRUE)
 }
x <- 1:10
plot(x,conversion_rates,type="b",pch=16,col="green",ylim=c(0.2,0.8),main="Conversion Rates for Plays between 41-50 yard line",xlab="Yards to go",ylab="Conversion Rate")
points(x,conversion_rates1,type="b",pch=16,col="red")
legend(6,0.8,legend=c("Teams that won","Teams that lost"),col=c("green","red"),lty=1:2)

```


```{r}
field_position_groups <- list(
  "1-10" = goforit %>% filter(yardline_100 >= 1 & yardline_100 <= 10),
  "11-20" = goforit %>% filter(yardline_100 >= 11 & yardline_100 <= 20),
  "21-30" = goforit %>% filter(yardline_100 >= 21 & yardline_100 <= 30),
  "31-40" = goforit %>% filter(yardline_100 >= 31 & yardline_100 <= 40),
  "41-50" = goforit %>% filter(yardline_100 >= 41 & yardline_100 <= 50),
  "51-60" = goforit %>% filter(yardline_100 >= 51 & yardline_100 <= 60),
  "61-70" = goforit %>% filter(yardline_100 >= 61 & yardline_100 <= 70),
  "71-80" = goforit %>% filter(yardline_100 >= 71 & yardline_100 <= 80),
  "81-90" = goforit %>% filter(yardline_100 >= 81 & yardline_100 <= 90),
  "91-100" = goforit %>% filter(yardline_100 >= 91 & yardline_100 <= 100)
)
```


```{r}
conversion_rates <- sapply(field_position_groups, function(group) {
  mean(group$fourth_down_converted, na.rm = TRUE)
})

positions <- names(conversion_rates)
rates <- conversion_rates

barplot(rates, names.arg = positions, main = "Fourth Down Conversion Rates by Field Position",
        xlab = "Field Position (yards from end zone)", ylab = "Conversion Rate", col = "green")

```

```{r}
for (position_group_name in names(field_position_groups)) {
  
  group_data <- field_position_groups[[position_group_name]]
  
  if (nrow(group_data) > 0) {
    
    conversion_rates <- numeric(10)
    
    for (yds in 1:10) {
      tbl <- group_data %>% filter(ydstogo == yds)
      conversion_rates[yds] <- mean(tbl$fourth_down_converted, na.rm = TRUE)
    }
    
    ydstogo <- 1:10
    correlation_value <- cor(ydstogo, conversion_rates, use = "complete.obs")
    
    print(paste("Correlation for field position group", position_group_name, ":", correlation_value))
    
    plot(1:10, conversion_rates, type = "b", pch = 16, col = "blue",
         main = paste("4th Down Conversion Rate for Field Position", position_group_name, "Yards from Endzone"),
         xlab = "Yards to Go", ylab = "Conversion Rate",
         ylim = c(0, 1), xlim = c(1, 10))
    
    Sys.sleep(2)
  } else {
    print(paste("No data available for field position group:", position_group_name))
  }
}
```

```{r}
group_data_11_20 <- field_position_groups[["11-20"]]

data_4th_and_2 <- group_data_11_20 %>% filter(ydstogo == 2)

num_data_points <- nrow(data_4th_and_2)

print(paste("Number of data points for 4th down and 2 yards to go in the '11-20' field position group:", num_data_points))
```
```{r}
goforit$play_type %>% table()
passes <- goforit %>% filter(play_type=="pass")
plot(passes$comp_air_epa,passes$comp_yac_epa)
goforit
```
```{r}
library(ggplot2)
library(gridExtra)
shortplays <-  goforit %>% filter(ydstogo <= 2)
par(mfrow=c(1,2))
plot1 <- ggplot(goforit, aes(x = play_type)) +
  geom_bar() +
  labs(title = "Fourth-Down Attempts Across Seasons")
plot2 <- ggplot(shortplays, aes(x = play_type)) +
  geom_bar() +
  labs(title = "Fourth and 1 and 2 ONLY")
grid.arrange(plot1, plot2, ncol=2)
```
```{r}
ggplot(goforit, aes(x = down, y = air_epa, color = as.factor(success))) +
  geom_point(position="jitter") +
  labs(title = "EPA on Fourth-Down Plays")
```

```{r}
for (i in 1:4){
goforit_40 <- goforit %>% filter(yardline_100 <= 50 & yardline_100 >= 40, qtr==i)
goforit_40_long <- goforit_40 %>% filter(ydstogo>=8)
goforit_40_mid <- goforit_40 %>% filter(ydstogo>2&ydstogo<8)
goforit_40_1 <- goforit_40 %>% filter(ydstogo<=2)
dens <- density(goforit_40_long$score_differential_post)
dens1 <- density(goforit_40_1$score_differential_post)
dens2 <- density(goforit_40_mid$score_differential_post)
plot(dens$x,dens$y,xlab="Posession Team Score - Defending Team Score",ylab="Density",ylim=c(0,0.065),col="red")
points(dens1$x,dens1$y,col="green")
points(dens2$x,dens2$y,col="yellow")
}
## Calculate expected value in each system and variance
mean(goforit_40_long$success)
mean(goforit_40_1$success)
counts <- c(nrow(goforit %>% filter(yardline_100 <= 50 & yardline_100 >= 40, qtr==1)),
nrow(goforit %>% filter(yardline_100 <=50 & yardline_100 >= 40, qtr==2)),
nrow(goforit %>% filter(yardline_100 <=50 & yardline_100 >= 40, qtr==3)),
nrow(goforit %>% filter(yardline_100 <=50 & yardline_100 >= 40, qtr==4)))
counts
```
> We can see that teams that are going for it are in general losing, especially when they are going for it with a large yards to go. The opposite actually happens when there is only 1 or 2 yards to go as the winning team is more likely to go for it than the losing team.
> The success rate also drops off by a lot when you go up in yards to go, which is very much expected as going for it from 1-2 yards gives you about a 60% chance to convert, but when it's 8+ yards it's a slim 25% chance.
> The 4 plots are the density curves for all 4 quarters, but we have to watch since there is a lot larger sample size in qtr 4 than any other quarter.
