---
title: "Team_Ranks"
output: md_document
date: "2024-11-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## R Markdown

```{r}
library(nflfastR)
data <- load_pbp(1999:2024)
dim(data)
head(data)
```

```{r}
allfours <- data %>% filter(down==4,yardline_100<=50,yardline_100>20)
kicks <- allfours %>% filter(play_type=="field_goal")
goforit <- allfours %>% filter(play_type!="field_goal") %>% filter(play_type!="punt") %>% filter(play_type!="no_play")
dens <- density(kicks$wpa)
plot(dens$x,dens$y,xlab="Win Probability Added",ylab="Density",main="Density of WPA on Attempted Field Goals")
abline(v=mean(kicks$wpa),col="red",pch=8)
```
```{r}
punts <- allfours %>% filter(play_type=="punt")
dens <- density(punts$wpa)
plot(dens$x,dens$y,xlab="Win Probability Added",ylab="Density",main="Density of WPA on Punts")
abline(v=mean(punts$wpa),col="red",pch=8)
```
```{r}
punt41 <- punts %>% filter(yardline_100==41)
dens <- density(punt41$epa)
plot(dens$x,dens$y,xlab="Expected Points Added",ylab="Density",main="Density of EPA on Punts")
abline(v=mean(punt41$wpa),col="red",pch=8)
msdpunt <- function(yard_line){
  new_df <- punts %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
msdkick <- function(yard_line){
  new_df <- kicks %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
msdgoforit1 <- function(yard_line){
  new_df <- goforit %>% filter(ydstogo==1) %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
msdgoforit23 <- function(yard_line){
  new_df <- goforit %>% filter(ydstogo==2|ydstogo==3) %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
msdgoforit47 <- function(yard_line){
  new_df <- goforit %>% filter(ydstogo>=4&ydstogo<=7) %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
msdgoforit8 <- function(yard_line){
  new_df <- goforit %>% filter(ydstogo>=8) %>% filter(yardline_100==yard_line)
  c(mean(new_df$epa),sd(new_df$epa))
}
```
```{r}
kick_epa_mean <- c()
punt_wpa_mean <- sapply(41:50,msdpunt)[1,]
kick_wpa_mean <- sapply(41:50,msdkick)[1,]
punt_wpa_sd <- sapply(41:50,msdpunt)[2,]
kick_wpa_sd <- sapply(41:50,msdkick)[2,]
goforit_wpa_mean1 <- sapply(41:50,msdgoforit1)[1,]
goforit_wpa_sd1 <- sapply(41:50,msdgoforit1)[2,]
goforit_wpa_mean2 <- sapply(41:50,msdgoforit23)[1,]
goforit_wpa_sd2 <- sapply(41:50,msdgoforit23)[2,]
goforit_wpa_mean4 <- sapply(41:50,msdgoforit47)[1,]
goforit_wpa_sd4 <- sapply(41:50,msdgoforit47)[2,]
goforit_wpa_mean8 <- sapply(41:50,msdgoforit8)[1,]
goforit_wpa_sd8 <- sapply(41:50,msdgoforit8)[2,]
df <- data.frame(punt_wpa_mean,kick_wpa_mean,goforit_wpa_mean1,goforit_wpa_mean2,goforit_wpa_mean4,goforit_wpa_mean8,punt_wpa_sd,kick_wpa_sd,goforit_wpa_sd1,goforit_wpa_sd2,goforit_wpa_sd4,goforit_wpa_sd8)
rownames(df) <- c("41","42","43","44","45","46","47","48","49","50")
colnames(df) <- c("Punts","Field Goals","4th and 1","4th and 2&3","4th and 4-7","4th and 8+","SD of Punts","SD of Field Goals","SD of 4th and 1","SD of 4th and 2&3","SD of 4th and 4-7", "SD of 4th and 8+")
df
```


In order to build functions individualized for each team that outlines their decision in certain scenarios, the equation should look something like this:

> mu + k(sigma)

where mu and sigma are mean and standard deviation respectively and the k is an unknown coefficient.

Our goal is to find out the specific k coefficient for every combination of on-field factors. The factors we should probably consider are:
- Time Remaining in Game
- Teams Previous History
- Current Point Differential

k will be a positive number but we should make the median of k = 1.
- half_seconds_remaining will vary inversely with k
- game_seconds_remaining will also vary inversely with k
- score_differential will vary inversely with k

Also lets look into teams previous history

```{r}
individualteamprobs <- function(team){
  falcons <- allfours %>% filter(posteam==team)
  falcons$binydstogo <-  cut(falcons$ydstogo, breaks=c(0,1,3,7,50), labels=c("4th and 1","4th and 2-3","4th and 4-7","4th and 8+"))
  falconsgo <- falcons %>% filter(play_type!="field_goal") %>% filter(play_type!="punt") %>% filter(play_type!="no_play")
  falconskick <- falcons %>% filter(play_type=="field_goal")
  falconspunt <- falcons %>% filter(play_type=="punt")
  falconspunt$puntspot <- falconspunt$yardline_100 - falconspunt$kick_distance
  falconspunt$puntsuccess <- ifelse(falconspunt$puntspot <= 10, 1, 0)
  punttbl <- table(falconspunt$binydstogo,falconspunt$puntsuccess)
  kicktbl <- table(falconskick$binydstogo,falconskick$success)
  gotbl <- table(falconsgo$binydstogo,falconsgo$success)
  colnames(gotbl) <- c("Failed","Success")
  colnames(punttbl) <- c("Failed","Success")
  colnames(kicktbl) <- c("Failed","Success")
  govec <- c((gotbl[1,2]/sum(gotbl[1,])),(gotbl[2,2]/sum(gotbl[2,])),(gotbl[3,2]/sum(gotbl[3,])),(gotbl[4,2]/sum(gotbl[4,])))
  puntvec <- c((punttbl[1,2]/sum(punttbl[1,])),(punttbl[2,2]/sum(punttbl[2,])),(punttbl[3,2]/sum(punttbl[3,])),(punttbl[4,2]/sum(punttbl[4,])))
  kickvec <- c((kicktbl[1,2]/sum(kicktbl[1,])),(kicktbl[2,2]/sum(kicktbl[2,])),(kicktbl[3,2]/sum(kicktbl[3,])),(kicktbl[4,2]/sum(kicktbl[4,])))
  labs <- c("4th and 1","4th and 2-3","4th and 4-7","4th and 8+")
  probmatrix <- data.frame(labs,govec*100,kickvec*100,puntvec*100)
  colnames(probmatrix) <- c("Yards to go","Percent Chance of Conversion", "Percent Chance of Field Goal","Percent Chance of Punt inside 10 yard line")
  probmatrix
}
individualteamprobs("TEN")
```

```{r}
falcons <- allfours %>% filter(posteam=="PHI")
  falcons$binydstogo <-  cut(falcons$ydstogo, breaks=c(0,1,3,7,50), labels=c("4th and 1","4th and 2-3","4th and 4-7","4th and 8+"))
  falconsgo <- falcons %>% filter(play_type!="field_goal") %>% filter(play_type!="punt") %>% filter(play_type!="no_play")
  falconskick <- falcons %>% filter(play_type=="field_goal")
  falconspunt <- falcons %>% filter(play_type=="punt")
  falconspunt$puntspot <- falconspunt$yardline_100 - falconspunt$kick_distance
  falconspunt$puntsuccess <- ifelse(falconspunt$puntspot <= 10, 1, 0)
  punttbl <- table(falconspunt$binydstogo,falconspunt$puntsuccess)
  kicktbl <- table(falconskick$binydstogo,falconskick$success)
  gotbl <- table(falconsgo$binydstogo,falconsgo$success)
  colnames(gotbl) <- c("Failed","Success")
  colnames(punttbl) <- c("Failed","Success")
  colnames(kicktbl) <- c("Failed","Success")
  govec <- c((gotbl[1,2]/sum(gotbl[1,])),(gotbl[2,2]/sum(gotbl[2,])),(gotbl[3,2]/sum(gotbl[3,])),(gotbl[4,2]/sum(gotbl[4,])))
  puntvec <- c((punttbl[1,2]/sum(punttbl[1,])),(punttbl[2,2]/sum(punttbl[2,])),(punttbl[3,2]/sum(punttbl[3,])),(punttbl[4,2]/sum(punttbl[4,])))
  kickvec <- c((kicktbl[1,2]/sum(kicktbl[1,])),(kicktbl[2,2]/sum(kicktbl[2,])),(kicktbl[3,2]/sum(kicktbl[3,])),(kicktbl[4,2]/sum(kicktbl[4,])))
  labs <- c("4th and 1","4th and 2-3","4th and 4-7","4th and 8+")
  probmatrix <- data.frame(labs,govec*100,kickvec*100,puntvec*100)
  colnames(probmatrix) <- c("Yards to go","Percent Chance of Conversion", "Percent Chance of Field Goal","Percent Chance of Punt inside 10 yard line")
falcons
```
## Creation of Team's History Variables

We want to divide up team's history in order to incorporate as many variables as possible to achieve a better k.
- Aggressiveness (Attempts)
- Success Rates

## ACF Plot of Conversion Rates

Here we are looking to see if the conversion success rates for individual teams are correlated with lags from previous years. We go back to 1999 in this data set.

```{r}
new_vec <- c()
falcons <- allfours %>% filter(posteam=="MIN")
for (i in 1999:2024){
  filtereddata1 <- falcons %>% filter(play_type=="run"|play_type=="pass") %>% filter(game_date>=i&game_date<i+1)
  new_vec[i-1998] <- filtereddata1$success %>% mean()
  new_vec <- na.omit(new_vec)
}
new_df <- data.frame(new_vec,lagged=lag(new_vec))
acf(ts(new_vec),max.lag=20)
```

After looking at a lot of different teams, there have been no real significant autocorrelations in the lag plot for the mean conversion rates over the past year.
